---
title: "Instability as a Data Quality Issue in Wikipedia"
subtitle: "*Evidence from Diversity-Related Topics in English Wikipedia*"
license: "CC BY"
date: 2026-01-01
date-format: "DD.MM.YYYY"
date-modified: last-modified
keywords:
  - Wikipedia
  - data quality
  - instability
  - diversity
  - edit conflict
  - burstiness
author:
  - name: Anju Yamada
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    toc-title: "Contents"
    embed-resources: true
    code-fold: true
    code-tools: false
    code-summary: "Show code"
    theme: litera
    fig-width: 10
    fig-height: 7
    fig-dpi: 300
    number-sections: true
    bibliography: references.bib
  
---
## Introduction

Wikipedia is one of the most widely used knowledge infrastructures in the world. However, its content is not fixed. It is continuously updated, revised, and contested by editors. This dynamic character makes Wikipedia not only a source of information, but also a data system shaped by ongoing social processes. Early work comparing Wikipedia with traditional encyclopedias has shown that its accuracy can be comparable in many cases, while still varying across articles and topics [@giles_special_2005].

Prior research has examined the quality and reliability of Wikipedia articles, emphasizing the implications of its open and collaborative nature. Because Wikipedia allows a large number of contributors to continuously edit and revise content, article quality is inherently dynamic and subject to variation over time [@moas_automatic_2024]. In addition to textual characteristics, revision history and editing behavior have been identified as important indicators for evaluating article quality, reflecting the process through which content is produced and maintained [@raman_classifying_2020; @zeng_computing_2006].

At the same time, editing activity in Wikipedia is shaped by social interaction and conflict. Previous studies have shown that intense disagreements, often manifested as mutual reverts, characterize a subset of highly contested articles [@sumi_edit_2011]. These findings suggest that certain topics may exhibit higher levels of revert-related instability, reflecting ongoing editorial contestation.

In parallel, research on human dynamics demonstrates that activity is often temporally concentrated rather than evenly distributed. Human activity patterns have been shown to follow bursty dynamics, where short periods of intense activity are followed by longer periods of inactivity [@barabasi_origin_2005]. Editing behavior in Wikipedia exhibits similar bursty patterns, where periods of intense activity are followed by relative inactivity [@choi_individual-driven_2021]. Such burstiness may reflect external attention, coordinated interaction, or event-driven editing processes, indicating that instability can also emerge through temporal concentration rather than explicit conflict.

Furthermore, prior work has identified systematic biases in Wikipedia content, particularly in relation to gender and diversity. Studies show that women are underrepresented and differently characterized in Wikipedia articles [@reagle_gender_2011; @graells-garrido_first_2015; @sun_men_2021]. These findings suggest that diversity-related topics are shaped by distinct social and editorial dynamics, including uneven attention and potentially heightened contestation.

Despite these advances, existing research has primarily focused on content-based quality measures or revision-based indicators such as edit counts and revert rates. Less attention has been given to the temporal stability of articles as a dimension of data quality. In particular, it remains unclear whether certain types of topics—such as diversity-related pages—are systematically more unstable, and whether such instability affects their reliability as data.

This report addresses this gap by conceptualizing instability as a data quality issue. Rather than focusing only on the volume of edits or the presence of conflict, it examines how editing activity is distributed over time. Specifically, it investigates whether diversity-related Wikipedia pages differ from other pages in terms of revert-related instability, bursty editing dynamics, and the concentration of edits across time.

This report focuses on English Wikipedia and examines whether diversity-related pages differ from other pages in terms of revert-related instability, bursty editing dynamics, and the concentration of edits across time. English Wikipedia is used as the empirical setting because it is the largest and most extensively studied language edition, providing a rich and well-documented dataset for analyzing editing behavior and data quality. Prior research has also predominantly relied on English Wikipedia when examining revision-based indicators of quality and collaboration dynamics, making it a suitable and comparable context for this study [@moas_automatic_2024; @raman_classifying_2020].

**Research Questions**

- Are diversity-related Wikipedia pages more unstable than other pages?
- Does instability appear primarily as revert-related conflict or as bursty temporal concentration?
- Can mitigation strategies improve data reliability?

**Hypotheses**

- **H1**: Diversity-related pages show stronger editing bursts.
- **H2**: Diversity-related pages are more likely to experience burst days.
- **H3**: Diversity-related pages have higher revert-related instability.

## Data and Research Design

This report uses Wikimedia data for English Wikipedia, combining page-level metadata, daily pageviews, and edit logs. Diversity-related pages are defined conservatively using the label `Culture.Biography.Women`. Biography pages are also identified using `Culture.Biography.Biography*`.

Our analysis distinguishes between three related dimensions of instability:

1. **Structural instability** at the page level  
   This captures whether editing activity is concentrated into a small number of days for a page overall.

2. **Temporal instability** at the day level  
   This captures whether a specific page-day is unusually intense relative to that page’s normal editing pattern.

3. **Revert-related instability**  
   This captures the extent to which edits are undone, reflecting disagreement and conflict in editorial activity.

This report interprets all three as potential data quality issues, especially in relation to consistency and reliability.


```{r}
#| label: 00_setup
#| include: false
#| message: false
#| warning: false
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
options(repos = c(CRAN = "https://cran.r-project.org"))
set.seed(2026)

library(jsonlite)
library(dplyr)
library(purrr)
library(stringr)
library(lubridate)
library(tidyr)
library(ggplot2)
library(readr)
library(knitr)
library(kableExtra)
library(scales)
library(forcats)
```

```{r, warning=FALSE, message=FALSE}
#| include: false
#| message: false
#| warning: false
output_dir <- "output"
fig_dir <- file.path(output_dir, "figures")
table_dir <- file.path(output_dir, "tables")

if (!dir.exists(output_dir)) dir.create(output_dir)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
if (!dir.exists(table_dir)) dir.create(table_dir, recursive = TRUE)

knitr::opts_chunk$set(
  fig.path = "output/figures/"
)

save_table_csv <- function(x, filename) {
  readr::write_csv(x, file.path(table_dir, filename))
}

save_plot_png <- function(plot_obj, filename, width = 8, height = 5, dpi = 300) {
  ggplot2::ggsave(
    filename = file.path(fig_dir, filename),
    plot = plot_obj,
    width = width,
    height = height,
    dpi = dpi
  )
}
```

```{r, warning=FALSE, message=FALSE}
#| include: false
#| message: false
#| warning: false
topic_colors <- c(
  "Diversity-related pages" = "#7B61FF",
  "Other pages" = "#2BB3B1"
)

design_colors <- c(
  "Raw" = "#4C78A8",
  "Revert filtered" = "#F58518",
  "Revert weighted" = "#54A24B",
  "Burst weighted" = "#E45756",
  "Event-day trimmed" = "#B279A2",
  "Combined" = "#72B7B2"
)

theme_report <- function() {
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 17),
      plot.subtitle = element_text(size = 12, color = "gray30"),
      plot.caption = element_text(size = 10, color = "gray40"),
      axis.title = element_text(face = "bold", size = 12),
      axis.text.x = element_text(color = "gray20", size = 11),
      axis.text.y = element_text(color = "gray20", size = 11),
      legend.text = element_text(size = 11),
      panel.grid.minor = element_blank(),
      legend.position = "top",
      legend.title = element_blank(),
      plot.title.position = "plot"
    )
}
```

```{r, warning=FALSE, message=FALSE}

#| label: 01_import_data
#| include: false
#| warning: false
#| message: false

n_read <- 50000

pageviews_path <- file.path("Data", "page_views.json.gz")
pageinfo_path  <- file.path("Data", "page_info.json.gz")
enwiki_path    <- file.path("Data", "enwiki.json.gz")

print(pageviews_path)
print(pageinfo_path)
print(enwiki_path)

stopifnot(file.exists(pageviews_path))
stopifnot(file.exists(pageinfo_path))
stopifnot(file.exists(enwiki_path))

con_pageviews <- textConnection(readLines(gzfile(pageviews_path), n = n_read))
pageviews_sample <- stream_in(con_pageviews, verbose = FALSE)
close(con_pageviews)

con_pageinfo <- textConnection(readLines(gzfile(pageinfo_path), n = n_read))
pageinfo_sample <- stream_in(con_pageinfo, verbose = FALSE)
close(con_pageinfo)

con_enwiki <- textConnection(readLines(gzfile(enwiki_path), n = n_read))
enwiki_sample <- stream_in(con_enwiki, verbose = FALSE)
close(con_enwiki)
```


```{r, warning=FALSE, message=FALSE}
#| label: 02_helpers
#| include: false
#| message: false
#| warning: false
get_label_prob <- function(label_df, target_label) {
  match <- label_df %>% filter(label == target_label)
  if (nrow(match) == 0) return(0)
  match$probability[1]
}

has_any_tag <- function(tags, target_tags) {
  if (is.null(tags) || length(tags) == 0) return(FALSE)
  if (is.list(tags) && !is.character(tags)) {
    tags <- unlist(tags, use.names = FALSE)
  }
  if (length(tags) == 1 && is.na(tags)) return(FALSE)
  tags_chr <- as.character(tags)
  any(tags_chr %in% target_tags)
}

safe_sd <- function(x) {
  if (length(x) <= 1) return(NA_real_)
  sd(x, na.rm = TRUE)
}

safe_quantile <- function(x, p) {
  if (length(x) == 0 || all(is.na(x))) return(NA_real_)
  as.numeric(quantile(x, p, na.rm = TRUE, type = 7))
}

```


```{r, warning=FALSE, message=FALSE}
#| include: false
#| message: false
#| warning: false
#| label: 03_prepare_data

pv_en <- pageviews_sample %>%
  filter(wiki_db == "enwiki")

pi_en <- pageinfo_sample %>%
  filter(wiki_db == "enwiki") %>%
  mutate(
    women_prob = map_dbl(predicted_labels, ~ get_label_prob(.x, "Culture.Biography.Women")),
    bio_prob   = map_dbl(predicted_labels, ~ get_label_prob(.x, "Culture.Biography.Biography*")),
    women_topic = women_prob > 0.20,
    biography_topic = bio_prob > 0.50,
    diversity_topic = women_topic
  )

pi_en_small <- pi_en %>%
  select(
    page_id,
    page_title,
    qid,
    total_pageviews = pageviews,
    women_prob,
    bio_prob,
    women_topic,
    biography_topic,
    diversity_topic
  )

edits_en <- enwiki_sample %>%
  filter(!is_bot) %>%
  mutate(
    revision_time = ymd_hms(revision_timestamp, quiet = TRUE),
    edit_day = as.Date(revision_time),
    is_reverted = map_lgl(revision_tags, ~ has_any_tag(.x, c("mw-reverted"))),
    is_revert_action = map_lgl(revision_tags, ~ has_any_tag(.x, c("mw-undo", "mw-rollback", "mw-manual-revert")))
  ) %>%
  left_join(pi_en_small, by = "page_id") %>%
  filter(!is.na(diversity_topic), !is.na(edit_day))

```


```{r, warning=FALSE, message=FALSE}
#| include: false
#| message: false
#| warning: false
#| label: 04_daily_analysis

pv_date_col <- intersect(names(pv_en), c("date", "timestamp", "view_day", "day"))

pv_daily_en <- pv_en %>%
  mutate(
    view_day_raw = .data[[pv_date_col[1]]],
    view_day = as.Date(view_day_raw)
  ) %>%
  group_by(page_id, view_day) %>%
  summarise(
    pageviews_day = sum(pageviews, na.rm = TRUE),
    .groups = "drop"
  )

daily_edits <- edits_en %>%
  group_by(page_id, page_title, diversity_topic, biography_topic, edit_day) %>%
  summarise(
    edits_day = n(),
    editors_day = n_distinct(user_text),
    reverted_day = sum(is_reverted, na.rm = TRUE),
    revert_action_day = sum(is_revert_action, na.rm = TRUE),
    reverted_share_day = reverted_day / edits_day,
    revert_action_share_day = revert_action_day / edits_day,
    .groups = "drop"
  )

daily_analysis <- daily_edits %>%
  group_by(page_id) %>%
  mutate(
    active_days = n(),
    mean_edits_page = mean(edits_day, na.rm = TRUE),
    sd_edits_page = safe_sd(edits_day),
    p95_edits_page = safe_quantile(edits_day, 0.95),
    max_edits_page = max(edits_day, na.rm = TRUE),
    burst_day = case_when(
      active_days < 5 ~ FALSE,
      TRUE ~ edits_day > p95_edits_page
    ),
    burst_day_sd = case_when(
      is.na(sd_edits_page) ~ FALSE,
      TRUE ~ edits_day > (mean_edits_page + 2 * sd_edits_page)
    ),
    burst_intensity = edits_day / sum(edits_day, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  left_join(
    pv_daily_en,
    by = c("page_id" = "page_id", "edit_day" = "view_day")
  )

page_instability <- daily_analysis %>%
  group_by(page_id, page_title, diversity_topic, biography_topic) %>%
  summarise(
    n_active_days = n(),
    n_edits = sum(edits_day, na.rm = TRUE),
    mean_edits_day = mean(edits_day, na.rm = TRUE),
    max_edits_day = max(edits_day, na.rm = TRUE),
    mean_editors_day = mean(editors_day, na.rm = TRUE),
    mean_reverted_share_day = mean(reverted_share_day, na.rm = TRUE),
    burst_day_rate = mean(burst_day, na.rm = TRUE),
    edit_burst = max(edits_day, na.rm = TRUE) / sum(edits_day, na.rm = TRUE),
    mean_pageviews_day = mean(pageviews_day, na.rm = TRUE),
    .groups = "drop"
  )
```
## Structural Instability at the Page Level

This report first examines structural instability at the page level. Here, instability is defined as the concentration of edits within a small number of days. This captures whether some pages are structurally more burst-prone than others. Table 1 reports descriptive statistics of page-level instability by topic group.

```{r, warning=FALSE, message=FALSE}
#| message: false
#| warning: false
#| label: tbl-page-results-table
#| tbl-cap: "Page-level instability by topic group"

result_pages_all <- page_instability %>%
  group_by(diversity_topic) %>%
  summarise(
    n_pages = n(),
    mean_edit_burst = mean(edit_burst, na.rm = TRUE),
    median_edit_burst = median(edit_burst, na.rm = TRUE),
    mean_burst_day_rate = mean(burst_day_rate, na.rm = TRUE),
    mean_reverted_share_day = mean(mean_reverted_share_day, na.rm = TRUE),
    mean_n_edits = mean(n_edits, na.rm = TRUE),
    mean_n_active_days = mean(n_active_days, na.rm = TRUE),
    .groups = "drop"
  )

save_table_csv(result_pages_all, "result_pages_all.csv")

result_pages_all |>
  kable(digits = 3, caption = "Page-level instability by topic group") |>
  kable_styling(full_width = FALSE)

```
Diversity-related pages show a higher mean edit burst (0.58) compared to other pages (0.44), indicating stronger concentration of edits.

Figure 1 visualizes the distribution of page-level edit burst by topic group.
```{r, warning=FALSE, message=FALSE}
#| message: false
#| warning: false
#| label: fig-page-results-plot
#| fig-cap: "Structural editing burstiness by topic"

p_page_burst <- page_instability %>%
  mutate(
    topic_label = if_else(
      diversity_topic,
      "Diversity-related pages",
      "Other pages"
    )
  ) %>%
  ggplot(aes(x = topic_label, y = edit_burst, fill = topic_label)) +
  geom_boxplot(width = 0.65, alpha = 0.85, outlier.alpha = 0.25) +
  stat_summary(
    fun = median,
    geom = "point",
    shape = 23,
    size = 3,
    fill = "white",
    color = "black"
  ) +
  scale_fill_manual(values = topic_colors) +
  scale_y_continuous(
    labels = label_percent(accuracy = 1),
    expand = expansion(mult = c(0.02, 0.08))
  ) +
  labs(
    title = "Structural editing burstiness is higher for diversity-related pages",
    subtitle = "Page-level edit concentration, measured as the share of all edits occurring on the single most active day",
    x = NULL,
    y = "Edit concentration within page (%)",
    caption = "White diamond marks the median"
  ) +
  theme_report()

save_plot_png(p_page_burst, "structural_edit_burst_by_topic.png")
p_page_burst

```
Diversity-related pages show consistently higher burst values, indicating stronger concentration of edits within a small number of days.

```{r, warning=FALSE, message=FALSE}
#| label: 05_hypothesis_tests_page_burst
#| message: false
#| warning: false

test_h1_all <- t.test(edit_burst ~ diversity_topic, data = page_instability)

```

A Welch two-sample t-test shows that diversity-related pages exhibit significantly higher structural burstiness ($t = -8.41$, $df = 552.99$, $p < .001$). The mean burstiness is 0.580 for diversity-related pages and 0.440 for non-diversity pages.

This indicates that diversity-related pages show stronger temporal concentration of editing activity at the page level. **Thus, H1 is strongly supported.**
```{r, warning=FALSE, message=FALSE}

```

## Temporal Instability at the Day Level

This report next examines instability at the page-day level. Here, the goal is to identify whether certain days are unusually intense relative to a page’s typical editing pattern. Table 2 reports day-level instability measures by topic group.

```{r, warning=FALSE, message=FALSE}
#| message: false
#| warning: false
#| label: tbl-day-results
#| tbl-cap: "Day-level instability by topic group"

result_days_all <- daily_analysis %>%
  group_by(diversity_topic) %>%
  summarise(
    n_page_days = n(),
    n_pages = n_distinct(page_id),
    mean_edits_day = mean(edits_day, na.rm = TRUE),
    median_edits_day = median(edits_day, na.rm = TRUE),
    mean_editors_day = mean(editors_day, na.rm = TRUE),
    mean_reverted_share_day = mean(reverted_share_day, na.rm = TRUE),
    burst_day_rate = mean(burst_day, na.rm = TRUE),
    burst_day_rate_sd = mean(burst_day_sd, na.rm = TRUE),
    .groups = "drop"
  )

save_table_csv(result_days_all, "result_days_all.csv")

result_days_all |>
  kable(digits = 3, caption = "Day-level instability by topic group") |>
  kable_styling(full_width = FALSE)
```
The share of burst days is lower for diversity-related pages (0.019) than for other pages (0.031), suggesting that burst events are not more frequent in diversity-related topics.

Figure 2 shows the share of burst days by topic group.

```{r, warning=FALSE, message=FALSE}
#| label: fig-burst-day
#| fig-cap: "Share of burst days by topic"

burst_plot_data <- daily_analysis %>%
  group_by(diversity_topic) %>%
  summarise(
    burst_day_rate = mean(burst_day, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    topic_label = if_else(
      diversity_topic,
      "Diversity-related pages",
      "Other pages"
    )
  )

p_burst_day <- burst_plot_data %>%
  ggplot(aes(x = topic_label, y = burst_day_rate, fill = topic_label)) +
  geom_col(width = 0.62, alpha = 0.9) +
  geom_text(
    aes(label = label_percent(accuracy = 0.1)(burst_day_rate)),
    vjust = -0.35,
    size = 4
  ) +
  scale_fill_manual(values = topic_colors) +
  scale_y_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, max(burst_plot_data$burst_day_rate) * 1.18)
  ) +
  labs(
    title = "Burst days are not more common among diversity-related pages",
    subtitle = "Share of active page-days classified as burst events",
    x = NULL,
    y = "Burst days among active page-days (%)"
  ) +
  theme_report() +
  theme(legend.position = "none")

save_plot_png(p_burst_day, "burst_day_rate_by_topic.png")
p_burst_day
```
The share of burst days is lower for diversity-related pages, suggesting that burst events are not more frequent despite higher page-level burstiness.

```{r, warning=FALSE, message=FALSE}
#| message: false
#| warning: false
#| label: 06_hypothesis_tests_day

test_h2_all <- t.test(as.numeric(burst_day) ~ diversity_topic, data = daily_analysis)

```

The difference between diversity-related and non-diversity pages is statistically significant ($t = 3.03$, $df = 1814.2$, $p = .002$), but the direction runs counter to expectations. Non-diversity pages show a higher mean share of burst days (0.031) compared to diversity-related pages (0.019).

This indicates that diversity-related pages do not experience burst events more frequently, despite showing higher structural burstiness at the page level. **Thus, H2 is not supported.**

## Revert-related Instability

This section examines whether diversity-related pages show higher levels of revert-related instability. Here, instability is defined as the extent to which edits are reverted, reflecting disagreement and conflict in editorial activity.

Figure 3 shows revert-related instability by topic group.
```{r, warning=FALSE, message=FALSE}
#| label: fig-revert-topic
#| fig-cap: "Revert-related instability by topic"

revert_plot_data <- page_instability %>%
  mutate(
    topic_label = if_else(
      diversity_topic,
      "Diversity-related pages",
      "Other pages"
    )
  )

p_revert_topic <- revert_plot_data %>%
  ggplot(aes(x = topic_label, y = mean_reverted_share_day, fill = topic_label)) +
  geom_boxplot(width = 0.65, alpha = 0.85, outlier.alpha = 0.25) +
  stat_summary(
    fun = median,
    geom = "point",
    shape = 23,
    size = 3,
    fill = "white",
    color = "black"
  ) +
  scale_fill_manual(values = topic_colors) +
  scale_y_continuous(
    labels = label_percent(accuracy = 1),
    expand = expansion(mult = c(0.02, 0.08))
  ) +
  labs(
    title = "Revert-related instability does not differ by topic group",
    subtitle = "Page-level reverted-edit share across diversity-related and other pages",
    x = NULL,
    y = "Mean reverted-edit share (%)",
    caption = "White diamond marks the median"
  ) +
  theme_report() +
  theme(legend.position = "none")

save_plot_png(p_revert_topic, "revert_instability_by_topic.png")
p_revert_topic
```

```{r, warning=FALSE, message=FALSE}
#| label: 07_hypothesis_tests_revert
#| message: false
#| warning: false

test_h3_all <- t.test(mean_reverted_share_day ~ diversity_topic, data = page_instability)
```
A Welch two-sample t-test shows no statistically significant difference in the mean share of reverted edits between diversity-related and non-diversity pages ($t = 0.36$, $df = 540.6$, $p = .716$). The mean reverted share is 0.179 for non-diversity pages and 0.174 for diversity-related pages.

This indicates that diversity-related pages are not more revert-prone than other pages. **Thus, H3 is not supported.**

## Mitigation Strategies

This study evaluates several mitigation strategies to assess whether data reliability can be improved once instability has been detected.

- revert-based filtering
- revert-based weighting
- topic/burst-based weighting
- event-day trimming
- combined mitigation

```{r}
#| label: 08_mitigation_prep
#| warning: false
#| message: false

revert_threshold <- quantile(page_instability$mean_reverted_share_day, 0.95, na.rm = TRUE)

page_revert_filtered <- page_instability %>%
  filter(mean_reverted_share_day <= revert_threshold)

page_revert_weighted <- page_instability %>%
  mutate(weight_revert = 1 / (1 + mean_reverted_share_day))

page_burst_weighted <- page_instability %>%
  mutate(weight_burst = 1 / (1 + edit_burst))

daily_trimmed <- daily_analysis %>%
  filter(!burst_day)

daily_trimmed_page <- daily_trimmed %>%
  group_by(page_id, page_title, diversity_topic, biography_topic) %>%
  summarise(
    n_active_days = n(),
    n_edits = sum(edits_day, na.rm = TRUE),
    mean_reverted_share_day = mean(reverted_share_day, na.rm = TRUE),
    edit_burst = max(edits_day, na.rm = TRUE) / sum(edits_day, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(weight_combined = 1 / (1 + edit_burst))
```
Table 3 compares instability measures across mitigation strategies.
```{r, warning=FALSE, message=FALSE}
#| label: tbl-mitigation
#| tbl-cap: "Mitigation comparison across designs"
#| message: false
#| warning: false

mitigation_compare <- bind_rows(
  page_instability %>%
    group_by(diversity_topic) %>%
    summarise(
      design = "raw",
      mean_reverted_share_day = mean(mean_reverted_share_day, na.rm = TRUE),
      mean_edit_burst = mean(edit_burst, na.rm = TRUE),
      mean_n_edits = mean(n_edits, na.rm = TRUE),
      .groups = "drop"
    ),
  page_revert_filtered %>%
    group_by(diversity_topic) %>%
    summarise(
      design = "revert_filtered",
      mean_reverted_share_day = mean(mean_reverted_share_day, na.rm = TRUE),
      mean_edit_burst = mean(edit_burst, na.rm = TRUE),
      mean_n_edits = mean(n_edits, na.rm = TRUE),
      .groups = "drop"
    ),
  page_revert_weighted %>%
    group_by(diversity_topic) %>%
    summarise(
      design = "revert_weighted",
      mean_reverted_share_day = weighted.mean(mean_reverted_share_day, weight_revert, na.rm = TRUE),
      mean_edit_burst = weighted.mean(edit_burst, weight_revert, na.rm = TRUE),
      mean_n_edits = weighted.mean(n_edits, weight_revert, na.rm = TRUE),
      .groups = "drop"
    ),
  page_burst_weighted %>%
    group_by(diversity_topic) %>%
    summarise(
      design = "topic_burst_weighted",
      mean_reverted_share_day = weighted.mean(mean_reverted_share_day, weight_burst, na.rm = TRUE),
      mean_edit_burst = weighted.mean(edit_burst, weight_burst, na.rm = TRUE),
      mean_n_edits = weighted.mean(n_edits, weight_burst, na.rm = TRUE),
      .groups = "drop"
    ),
  daily_trimmed %>%
    group_by(diversity_topic) %>%
    summarise(
      design = "event_day_trimmed",
      mean_reverted_share_day = mean(reverted_share_day, na.rm = TRUE),
      mean_edit_burst = mean(burst_intensity, na.rm = TRUE),
      mean_n_edits = mean(edits_day, na.rm = TRUE),
      .groups = "drop"
    ),
  daily_trimmed_page %>%
    group_by(diversity_topic) %>%
    summarise(
      design = "combined",
      mean_reverted_share_day = weighted.mean(mean_reverted_share_day, weight_combined, na.rm = TRUE),
      mean_edit_burst = weighted.mean(edit_burst, weight_combined, na.rm = TRUE),
      mean_n_edits = weighted.mean(n_edits, weight_combined, na.rm = TRUE),
      .groups = "drop"
    )
)

mitigation_compare_pretty <- mitigation_compare %>%
  mutate(
    topic_label = if_else(
      diversity_topic,
      "Diversity-related pages",
      "Other pages"
    ),
    design = recode(
      design,
      "raw" = "Raw",
      "revert_filtered" = "Revert filtered",
      "revert_weighted" = "Revert weighted",
      "topic_burst_weighted" = "Burst weighted",
      "event_day_trimmed" = "Event-day trimmed",
      "combined" = "Combined"
    ),
    design = factor(
      design,
      levels = c(
        "Raw",
        "Revert filtered",
        "Revert weighted",
        "Burst weighted",
        "Event-day trimmed",
        "Combined"
      )
    )
  )

save_table_csv(mitigation_compare, "mitigation_compare.csv")

mitigation_compare |>
  kable(digits = 3, caption = "Mitigation comparison across designs") |>
  kable_styling(full_width = FALSE)
```
Revert-based methods reduce the share of reverted edits but have limited effect on edit burst. In contrast, burst-based weighting and event-day trimming substantially reduce burst-related instability.

Figure 4 compares burst-related instability across mitigation strategies.
```{r, warning=FALSE, message=FALSE}
#| label: fig-mit-burst
#| fig-cap: "Mitigation comparison: burst-related instability"
#| message: false
#| warning: false

p_mit_burst <- mitigation_compare_pretty %>%
  ggplot(aes(x = design, y = mean_edit_burst, color = topic_label, group = topic_label)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_text(
    aes(label = label_percent(accuracy = 0.1)(mean_edit_burst)),
    vjust = -0.7,
    size = 3,
    show.legend = FALSE
  ) +
  scale_color_manual(values = topic_colors) +
  scale_y_continuous(
    labels = label_percent(accuracy = 1),
    expand = expansion(mult = c(0.03, 0.15))
  ) +
  labs(
    title = "Burst-related instability decreases under burst-focused mitigation",
    subtitle = "Comparison of average burst measure across mitigation designs",
    x = "Mitigation strategy",
    y = "Mean burst measure (%)"
  ) +
  theme_report() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))

save_plot_png(p_mit_burst, "mitigation_burst.png")
p_mit_burst
```
Burst-based weighting and event-day trimming substantially reduce burst-related instability, while revert-based methods have limited impact.
Figure 5 compares revert-related instability across mitigation strategies.
```{r, warning=FALSE, message=FALSE}
#| label: fig-mit-revert
#| fig-cap: "Mitigation comparison: revert-related instability"
#| message: false
#| warning: false

p_mit_revert <- mitigation_compare_pretty %>%
  ggplot(aes(x = design, y = mean_reverted_share_day, color = topic_label, group = topic_label)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_text(
    aes(label = label_percent(accuracy = 0.1)(mean_reverted_share_day)),
    vjust = -0.7,
    size = 3,
    show.legend = FALSE
  ) +
  scale_color_manual(values = topic_colors) +
  scale_y_continuous(
    labels = label_percent(accuracy = 1),
    expand = expansion(mult = c(0.03, 0.15))
  ) +
  labs(
    title = "Revert-focused mitigation reduces revert-related instability",
    subtitle = "Comparison of average reverted-share measure across mitigation designs",
    x = "Mitigation strategy",
    y = "Mean reverted-edit share (%)"
  ) +
  theme_report() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))

save_plot_png(p_mit_revert, "mitigation_revert.png")
p_mit_revert
```
Revert-focused mitigation strategies reduce the share of reverted edits, whereas burst-focused methods have little effect on this dimension.
## Discussion

This study examined whether diversity-related Wikipedia pages are more unstable than other pages, and whether that instability can be understood as a data quality issue.

The results provide **dimension-specific evidence**.

First, diversity-related pages are significantly more bursty at the page level. **H1 is strongly supported.** This indicates that editing activity is more temporally concentrated for these pages, suggesting a form of structural, topic-related instability.

Second, the day-level analysis does not support **H2**. Diversity-related pages are not more likely to experience burst days under the present definition. This indicates that their higher instability does not arise from more frequent burst events, but rather from a stronger concentration of edits within a smaller number of days.

Third, we do not find evidence that diversity-related pages are more revert-prone. **H3 is therefore not supported.** This suggests that instability in these pages is not primarily driven by overt revert conflict.

### Dimensions of Instability

Taken together, these findings suggest that instability in Wikipedia has at least two distinct dimensions:

- **Structural instability**  
  reflected in the concentration of edits across time  

- **Event-based instability**  
  reflected in burst days and short-term editing spikes  

In addition, revert-related instability captures a distinct dimension of **editorial conflict**, reflected in the share of edits that are undone.

These dimensions are related but empirically distinct. Diversity-related pages appear especially prone to structural instability, but not to higher levels of conflict or burst events.

### Implications for Data Quality

From a data quality perspective, this is important because instability is not adequately captured by revert rate alone. A page may not be highly revert-prone but still be unstable if its content is disproportionately shaped by a few concentrated editing periods.

### Mitigation Implications

The mitigation analysis reinforces this interpretation.

- Revert-based mitigation reduces conflict-related noise, but has limited effect on burst-related instability  
- Burst-based weighting and event-day trimming effectively reduce temporal concentration  

This suggests that instability is **multi-dimensional** and requires **targeted mitigation strategies**, rather than a single uniform solution.

## References {.appendix}

::: {#refs}
:::

## Software & Session Info {.appendix}

```{r}

#| label: session_info
#| warning: false
#| message: false

sessioninfo::session_info(pkgs = "attached", include_base = TRUE)
```